---
title: "NOAA StormData Analysis"
author: "Chris Brofft"
date: "9/23/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Synopsis:
The following analysis was completed using the NOAA storm database from the years 1950 thru 2011.  The goal of the analysis was to understand the major types of servere events and the toll taken by humanity and economy.  The resulting processing creates 2 plots showing the top 3 servere event types for human impact and economic impact.

# Setting up the environment:
```{r}
library(knitr)
library(ggplot2)
library(dplyr)
library(plyr)
```


### Staging the files and setting the directory.  
###The file will be downloaded from the source and unzipped and saved as stormdata.csv
```{r}
setwd("~/RepDataCourseProject2")
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
downloadFile <- "./repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileURL, downloadFile, method = "curl")
stormdata <-read.csv("repdata%2Fdata%2FStormData.csv.bz2",header = TRUE, sep = ",")
unlink(downloadFile)
```
# Data Processing:
### Calculate deaths and injuries from storms
```{r}
stormdata$death_injuries<-stormdata$FATALITIES+stormdata$INJURIES
```

###Convert from factor to character for property and crop damage
```{r}
stormdata$PROPDMGEXP<-as.character(stormdata$PROPDMGEXP)
stormdata$CROPDMGEXP<-as.character(stormdata$CROPDMGEXP)
```
###### Property and crop damage units and respective column numbers
```{r}
propdamage<-grep("PROPDMGEXP",colnames(stormdata))
cropdamage<-grep("CROPDMGEXP",colnames(stormdata))
```
###Calculating the units of damage for crop and property
```{r}
stormdata[stormdata$PROPDMGEXP=="",propdamage]<-"0"
stormdata[stormdata$PROPDMGEXP=="-",propdamage]<-"0"
stormdata[stormdata$PROPDMGEXP=="?",propdamage]<-"0"
stormdata[stormdata$PROPDMGEXP=="+",propdamage]<-"0"
stormdata[stormdata$PROPDMGEXP=="1",propdamage]<-"0"
stormdata[stormdata$PROPDMGEXP=="2",propdamage]<-"0"
stormdata[stormdata$PROPDMGEXP=="3",propdamage]<-"0"
stormdata[stormdata$PROPDMGEXP=="4",propdamage]<-"0"
stormdata[stormdata$PROPDMGEXP=="5",propdamage]<-"0"
stormdata[stormdata$PROPDMGEXP=="6",propdamage]<-"0"
stormdata[stormdata$PROPDMGEXP=="7",propdamage]<-"0"
stormdata[stormdata$PROPDMGEXP=="8",propdamage]<-"0"
stormdata[stormdata$PROPDMGEXP=="h",propdamage]<-"100"
stormdata[stormdata$PROPDMGEXP=="H",propdamage]<-"100"
stormdata[stormdata$PROPDMGEXP=="K",propdamage]<-"1000"
stormdata[stormdata$PROPDMGEXP=="m",propdamage]<-"1000000"
stormdata[stormdata$PROPDMGEXP=="M",propdamage]<-"1000000"
stormdata[stormdata$PROPDMGEXP=="B",propdamage]<-"1000000000"

stormdata[stormdata$CROPDMGEXP=="",cropdamage]<-"0"
stormdata[stormdata$CROPDMGEXP=="?",cropdamage]<-"0"
stormdata[stormdata$CROPDMGEXP=="2",cropdamage]<-"0"
stormdata[stormdata$CROPDMGEXP=="k",cropdamage]<-"1000"
stormdata[stormdata$CROPDMGEXP=="K",cropdamage]<-"1000"
stormdata[stormdata$CROPDMGEXP=="m",cropdamage]<-"1000000"
stormdata[stormdata$CROPDMGEXP=="M",cropdamage]<-"1000000"
stormdata[stormdata$CROPDMGEXP=="B",cropdamage]<-"1000000000"
```
### Convert property and crop damages to numeric values now
```{r}
stormdata$PROPDMGEXP<-as.numeric(stormdata$PROPDMGEXP)
stormdata$CROPDMGEXP<-as.numeric(stormdata$CROPDMGEXP)
```
###Calculate property and crop loss values
```{r}
stormdata$propertyloss<-stormdata$PROPDMG*stormdata$PROPDMGEXP
stormdata$croploss<-stormdata$CROPDMG*stormdata$CROPDMGEXP
```
### Get the total economic loss
```{r}
stormdata$totaleconomicloss<-stormdata$propertyloss+stormdata$croploss
```
### Get column numbers for human loss 
```{r}
col1<-grep("EVTYPE",colnames(stormdata))
col2<-grep("death_injuries",colnames(stormdata))
```
### Create data set for human loss with omitting NA values
```{r}
humanloss<-stormdata[,c(col1,col2)]
humanloss<-humanloss[complete.cases(humanloss), ]
```

### Get ccolumn numbers for economic loss 
```{r}
col1<-grep("EVTYPE",colnames(stormdata))
col2<-grep("economicloss",colnames(stormdata))
```
### Create economic loss with NA Values omitted
```{r}
economicloss<-stormdata[,c(col1,col2)]
economicloss<-economicloss[complete.cases(economicloss), ]
```
### Aggregation of  total human loss
```{r}
humanlossclean<-aggregate(death_injuries~EVTYPE,humanloss,sum,na.rm=TRUE)
### Sort human losses data in decreasing order of losses
humanlossclean<-humanlossclean[order(humanlossclean$death_injuries,decreasing = TRUE),]
```
### Create data set of top 3 events
```{r}
humanlossfinal<-humanlossclean[1:3,]
```
### Aggregation of  total economic losses
```{r}
ecolossclean<-aggregate(totaleconomicloss~EVTYPE,economicloss,sum,na.rm=TRUE)
```
### Sort economic losses data in decreasing order of losses
```{r}
ecolossclean<-ecolossclean[order(ecolossclean$totaleconomicloss,decreasing = TRUE),]
```
### Create dataset for top 3 events
```{r}
ecolossfinal<-ecolossclean[1:3,]
```
# Results:
### Create plot with top 3 servere types and human loss
```{r}
humanplot<-ggplot(data=humanlossfinal,aes(x=EVTYPE,y=death_injuries))
humanplot+geom_bar(stat="identity", fill = "red", color="black")+xlab("Servere Event")+ylab("Injuries and Death")
```
### Create plot with top 3 servere types and economicloss
```{r}
economyplot<-ggplot(data=ecolossfinal,aes(x=EVTYPE,y=totaleconomicloss))
economyplot+geom_bar(stat="identity", fill = "yellow", color="black")+xlab("Servere Event")+ylab("Economic Loss ($)")+
  scale_y_continuous(name="Economic Loss ($)", labels = scales::comma)
```